---
title: "TCGAutils: Helper functions for working with TCGA datasets"
author: "Marcel Ramos & Levi Waldron"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  BiocStyle::html_document:
    number_sections: yes
    toc: true
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{TCGAutils Essentials}
  %\VignetteEncoding{UTF-8}
---

# Overview

The `TCGAutils` package completes a suite of Bioconductor packages for
convenient access, integration, and analysis of *The Cancer Genome Atlas*.
It includes:
    0.    helpers for working with TCGA through the Bioconductor packages
    `r Biocpkg("MultiAssayExperiment")` (for coordinated representation and
    manipulation of multi-omits experiments) and `r Biocpkg("curatedTCGAData")`,
    which provides unrestricted TCGA data as `MultiAssayExperiment` objects,
    0.    helpers for importing TCGA data as from flat data structures such as
    `data.frame` or `DataFrame` read from delimited data structures provided by
    the Broad Institute’s Firehose, GenomicDataCommons, and
    0.    functions for interpreting TCGA barcodes and for mapping between
    barcodes and Universally Unique Identifiers (UUIDs).

# Installation

```{r, eval=FALSE}
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("TCGAutils")
```

Required packages for this vignette:

```{r,include=TRUE,results="hide",message=FALSE,warning=FALSE}
library(TCGAutils)
library(curatedTCGAData)
library(MultiAssayExperiment)
library(RTCGAToolbox)
library(BiocFileCache)
library(rtracklayer)
library(R.utils)
```

# `curatedTCGAData` utility functions

Functions such as `getSubtypeMap` and `getClinicalNames` provide information
on data inside a `r Biocpkg("MultiAssayExperiment")` object downloaded from
`r Biocpkg("curatedTCGAData")`. `sampleTables` and `TCGAsplitAssays` support
useful operations on these `MultiAssayExperiment` objects.

## obtaining TCGA as `MultiAssayExperiment` objects from `curatedTCGAData`

For a list of all available data types, use `dry.run = TRUE` and an
asterisk `*` as the assay input value:

```{r}
curatedTCGAData("COAD", "*", dry.run = TRUE, version = "1.1.38")
```

In this example, we download part of the Colon Adenocarcinoma (COAD) dataset
using `curatedTCGAData` via `ExperimentHub`. This command will download
data types such as `CNASeq`, `Mutation`, etc.:

```{r,include=TRUE,results="hide",message=FALSE,warning=FALSE}
coad <- curatedTCGAData(
    diseaseCode = "COAD", assays = c("CNASeq", "Mutation", "miRNA*",
    "RNASeq2*", "mRNAArray", "Methyl*"), version = "1.1.38", dry.run = FALSE
)
```

## `sampleTables`: what sample types are present in the data?

The `sampleTables` function gives a tally of available
samples in the dataset based on the TCGA barcode information.

```{r}
sampleTables(coad)
```

For reference in interpreting the sample type codes, see the `sampleTypes`
table:

```{r}
data("sampleTypes")
head(sampleTypes)
```

## `TCGAsplitAssays`: separate the data from different tissue types

TCGA datasets include multiple -omics for solid tumors, adjacent normal
tissues, blood-derived cancers and normals, and other tissue types, which may
be mixed together in a single dataset. The `MultiAssayExperiment` object
generated here has one patient per row of its `colData`, but each patient may
have two or more -omics profiles by any assay, whether due to assaying of
different types of tissues or to technical replication. `TCGAsplitAssays`
separates profiles from different tissue types (such as tumor and adjacent
normal) into different assays of the `MultiAssayExperiment` by taking a vector
of sample codes, and partitioning the current assays into assays with an
appended sample code:

```{r}
(tnmae <- TCGAsplitAssays(coad, c("01", "11")))
```

The `r Biocpkg("MultiAssayExperiment")` package then provides functionality to
merge replicate profiles for a single patient (`mergeReplicates()`), which
would now be appropriate but would **not** have been appropriate before
splitting different tissue types into different assays, because that would
average measurements  from tumors and normal tissues.

`MultiAssayExperiment` also defines the `MatchedAssayExperiment` class, which
eliminates any profiles not present across all assays and ensures identical
ordering of profiles (columns) in each assay. In this example, it will match
tumors to adjacent normals in subsequent assays:

```{r}
(matchmae <- as(tnmae[, , c(4, 6, 7)], "MatchedAssayExperiment"))
```

Only about 12 participants have both a matched tumor and solid normal sample.

## `getSubtypeMap`: manually curated molecular subtypes

Per-tumor subtypes are saved in the `metadata` of the `colData`
slot of `MultiAssayExperiment` objects downloaded from `curatedTCGAData`.
These subtypes were manually curated from the supplemental tables of all
primary TCGA publications:

```{r}
getSubtypeMap(coad)
```

## `getClinicalNames`: key "level 4" clinical &  pathological data

The `curatedTCGAData` `colData` contain hundreds of columns, obtained from
merging all unrestricted levels of clinical, pathological, and biospecimen data.
This function provides the names of "level 4" clinical/pathological variables,
which are the only ones provided by most other TCGA analysis tools.
Users may then use these variable names for subsetting or analysis, and may
even want to subset the `colData` to only these commonly used variables.

```{r}
getClinicalNames("COAD")
```

*Warning*: some names may not exactly match the `colData` names in the object
due to differences in variable types. These variables are kept separate and
differentiated with `x` and `y`. For example, `vital_status` in this case
corresponds to two different variables obtained from the pipeline. One variable
is interger type and the other character:

```{r}
class(colData(coad)[["vital_status.x"]])
class(colData(coad)[["vital_status.y"]])

table(colData(coad)[["vital_status.x"]])
table(colData(coad)[["vital_status.y"]])
```

Such conflicts should be inspected in this manner, and conflicts resolved by
choosing the more complete variable, or by treating any conflicting values as
unknown ("NA").

# Converting Assays to SummarizedExperiment

This section gives an overview of the operations that can be performed on
a given set of metadata obtained particularly from data-rich objects such
as those obtained from `curatedTCGAData`. There are several operations that
work with microRNA, methylation, mutation, and assays that have gene symbol
annotations.

## `CpGtoRanges`

Using the methylation annotations in
`IlluminaHumanMethylation450kanno.ilmn12.hg19` and the `minfi` package, we
look up CpG probes and convert to genomic coordinates with `CpGtoRanges`.
The function provides two assays, one with mapped probes and the other with
unmapped probes. Excluding unmapped probes can be done by setting the
`unmapped` argument to `FALSE`. This will run for both types of methylation
data (27k and 450k).

```{r}
methcoad <- CpGtoRanges(coad)
```


## `mirToRanges`

microRNA assays obtained from `curatedTCGAData` have annotated sequences
that can be converted to genomic ranges using the `mirbase.db` package.
The function looks up all sequences and converts them to ('hg19') ranges.
For those rows that cannot be found, an 'unranged' assay is introduced
in the resulting MultiAssayExperiment object.

```{r}
mircoad <- mirToRanges(coad)
```

## `qreduceTCGA`

The `qreduceTCGA` function converts `RaggedExperiment` mutation data objects
to `RangedSummarizedExperiment` using `org.Hs.eg.db` and the `qreduceTCGA`
utility function from `RaggedExperiment` to summarize 'silent' and 'non-silent'
mutations based on a 'Variant_Classification' metadata column in the original
object.

It uses 'hg19' transcript database ('TxDb') package internally to summarize
regions using `qreduceAssay`. The current genome build ('hg18') in the data
must be translated to 'hg19'.

In this example, we first set the appropriate build name in the mutation
dataset `COAD_Mutation-20160128` according to the
[NCBI website](https://www.ncbi.nlm.nih.gov/assembly/GCF_000001405.12/)
and we then use `seqlevelsStyle` to match the `UCSC` style in the chain.

```{r}
rag <- "COAD_Mutation-20160128"
# add the appropriate genome annotation
genome(coad[[rag]]) <- "NCBI36"
# change the style to UCSC
seqlevelsStyle(rowRanges(coad[[rag]])) <- "UCSC"

# inspect changes
seqlevels(rowRanges(coad[[rag]]))
genome(coad[[rag]])
```

Now we use `liftOver` from `rtracklayer` to translate 'hg18' builds
to 'hg19' using the chain file obtained via `AnnotationHub`. We use a
query to find the file. You can also query with terms such as
`c("Homo sapiens", "chain", "hg18", "hg19")` in the query function. We are
specifically looking for the chain file "hg18ToHg19.over.chain.gz".

```{r,include=TRUE,results="hide",message=FALSE,warning=FALSE}
library(AnnotationHub)
ah <- AnnotationHub()
```

```{r}
query(ah, "hg18ToHg19.over.chain.gz")
chain <- ah[["AH14220"]]
ranges19 <- rtracklayer::liftOver(rowRanges(coad[[rag]]), chain)
```

__Note__. The same can be done to convert `hg19` to `hg38` (the same build that
the Genomic Data Commons uses) with the corresponding chain file as obtained
above.

This will give us a list of ranges, each element corresponding to a single row
in the `RaggedExperiment`. We remove rows that had no matches in the `liftOver`
process and replace the ranges in the original `RaggedExperiment` with the
replacement method. Finally, we put the `RaggedExperiment` object back into the
`MultiAssayExperiment`.

```{r}
re19 <- coad[[rag]][as.logical(lengths(ranges19))]
ranges19 <- unlist(ranges19)
genome(ranges19) <- "hg19"
rowRanges(re19) <- ranges19
# replacement
coad[["COAD_Mutation-20160128"]] <- re19
rowRanges(re19)
```

Now that we have matching builds, we can finally run the `qreduceTCGA` function.

```{r}
coad <- qreduceTCGA(coad, keep.assay = TRUE)
```

## `symbolsToRanges`

In the cases where row annotations indicate gene symbols, the `symbolsToRanges`
utility function converts genes to genomic ranges and replaces existing
assays with `RangedSummarizedExperiment` objects. Gene annotations are given
as 'hg19' genomic regions.

```{r}
symbolsToRanges(coad)
```


# Importing TCGA text data files to Bioconductor classes

A few functions in the package accept either files or classes such as
`data.frame` and `FirehoseGISTIC` as input and return standard Bioconductor
classes.

### Work around for long file names on Windows

Due to file name length, Windows may not be able to read / display all files.
The workaround uses the `fileNames` argument from a character vector of file
names and will convert them to TCGA barcodes.

```{r}
## Load example file found in package
pkgDir <- system.file("extdata", package = "TCGAutils", mustWork = TRUE)
exonFile <- list.files(pkgDir, pattern = "cation\\.txt$", full.names = TRUE)
exonFile

## We add the original file prefix to query for the UUID and get the
## TCGAbarcode
filePrefix <- "unc.edu.32741f9a-9fec-441f-96b4-e504e62c5362.1755371."

## Add actual file name manually
makeGRangesListFromExonFiles(exonFile, getBarcodes = FALSE,
    fileNames = paste0(filePrefix, basename(exonFile)))
```

**NOTE**. Querying the API for legacy file names is no longer supported. In
this example, we leave the file name in rather than translate the file name
to a TCGA barcode, i.e., `getBarcodes = FALSE`. However, `filenameToBarcode` is
still supported for other file types.

## `makeGRangesListFromCopyNumber`

Other processed, genomic range-based data from TCGA data can be imported using
`makeGRangesListFromCopyNumber`. This tab-delimited data file of copy number
alterations from *bladder urothelial carcinoma* (BLCA) was obtained from the
Genomic Data Commons and is included in `TCGAUtils` as an example:

```{r}
grlFile <- system.file("extdata", "blca_cnaseq.txt", package = "TCGAutils")
grl <- read.table(grlFile)
head(grl)

makeGRangesListFromCopyNumber(grl, split.field = "Sample")

makeGRangesListFromCopyNumber(grl, split.field = "Sample",
    keep.extra.columns = TRUE)
```

## `makeSummarizedExperimentFromGISTIC`

This function is only used for converting the `FirehoseGISTIC` class of the
`r Biocpkg("RTCGAToolbox")` package. It allows the user to obtain thresholded
by gene data, probabilities and peak regions.

```{r}
tempDIR <- tempdir()
co <- getFirehoseData("COAD", clinical = FALSE, GISTIC = TRUE,
    destdir = tempDIR)

selectType(co, "GISTIC")
class(selectType(co, "GISTIC"))

makeSummarizedExperimentFromGISTIC(co, "Peaks")
```

## `mergeColData`: expanding the `colData` of a `MultiAssayExperiment`

This function merges a `data.frame` or `DataFrame` into the
`colData` of an existing `MultiAssayExperiment` object. It will match
column names and row names to do a full merge of both data sets. This
convenience function can be used, for example, to add subtype information
available for a subset of patients to the `colData`. Here is a simplified
example of adding a column to the `colData` `DataFrame`:

```{r}
race_df <- DataFrame(race_f = factor(colData(coad)[["race"]]),
    row.names = rownames(colData(coad)))
mergeColData(coad, race_df)
```

# Translating and interpreting TCGA identifiers

## GDC Data Updates

Data Release ([version 32.0](https://docs.gdc.cancer.gov/Data/Release_Notes/Data_Release_Notes/#data-release-320))
changed the behavior of the ID translation functions. This is due to the
removal of files whose UUIDs were translated. These files have been replaced
with newer runs of the pipeline. The new UUIDs can be mapped to the old UUIDs
with project specific maps located on GitHub at
https://github.com/NCI-GDC/gdc-docs/tree/develop/docs/Data/Release_Notes/GCv36_Manifests
as well as with the history lookup API endpoint.

Note. The GDC API version
[v3.28.0](https://docs.gdc.cancer.gov/API/Release_Notes/API_Release_Notes/#v3280)
has deprecated the `/legacy/files` endpoint; therefore, legacy data is no
longer available.

### UUID History Lookup

Facilities in the GDC API allow one to look up old UUIDs and obtain new ones.
The `UUIDhistory` function returns a `data.frame` with old and new UUIDs for
a single UUID lookup. This will allow one to map old UUIDs with new ones in
order to run a translation query. Old UUIDs have been removed from the API and
can no longer be translated by the functions provided in this package.

```{r}
UUIDhistory("0001801b-54b0-4551-8d7a-d66fb59429bf")
```

## Translation

The TCGA project has generated massive amounts of data. Some data can be
obtained with **U**niversally **U**nique **ID**entifiers (**UUID**) and other
data with TCGA barcodes. The Genomic Data Commons provides a JSON API for
mapping between UUID and barcode, but it is difficult for many people to
understand. `TCGAutils` makes simple functions available for two-way
translation between vectors of these identifiers.

### TCGA barcode to UUID

Here we translate the first two TCGA barcodes of the previous copy-number
alterations dataset to UUID:

```{r}
(xbarcode <- head(colnames(coad)[["COAD_CNASeq-20160128_simplified"]], 4L))
barcodeToUUID(xbarcode)
```

### UUID to TCGA barcode

Here we have a known case UUID that we want to translate into a TCGA barcode.

```{r}
UUIDtoBarcode("ae55b2d3-62a1-419e-9f9a-5ddfac356db4", from_type = "case_id")
```

In cases where we want to translate a known file UUID to the associated TCGA
patient barcode, we can use `UUIDtoBarcode`.

```{r}
UUIDtoBarcode("b4bce3ff-7fdc-4849-880b-56f2b348ceac", from_type = "file_id")
```

Translating aliquot UUIDs is also possible by providing a known aliquot UUID to
the function and giving a `from_type`, "aliquot_ids":

```{r}
UUIDtoBarcode("d85d8a17-8aea-49d3-8a03-8f13141c163b", from_type = "aliquot_ids")
```

Additional UUIDs may be supported in future versions.

### UUID to UUID

We can also translate from file UUIDs to case UUIDs and vice versa as long as
we know the input type. We can use the case UUID from the previous example to
get the associated file UUIDs using `UUIDtoUUID`. Note that this translation
is a one to many relationship, thus yielding a `data.frame` of file UUIDs for a
single case UUID.

```{r}
head(UUIDtoUUID("ae55b2d3-62a1-419e-9f9a-5ddfac356db4", to_type = "file_id"))
```

One possible way to verify that file IDs are matching case UUIDS is to
browse to the Genomic Data Commons webpage with the specific file UUID.
Here we look at the first file UUID entry in the output `data.frame`:

https://portal.gdc.cancer.gov/files/0ff55a5e-6058-4e0b-9641-e3cb375ff214

In the page we check that the case UUID matches the input.

## Parsing TCGA barcodes

Several functions exist for working with TCGA barcodes, the main function being
`TCGAbarcode`. It takes a TCGA barcode and returns information about
participant, sample, and/or portion.

```{r}
## Return participant barcodes
TCGAbarcode(xbarcode, participant = TRUE)

## Just return samples
TCGAbarcode(xbarcode, participant = FALSE, sample = TRUE)

## Include sample data as well
TCGAbarcode(xbarcode, participant = TRUE, sample = TRUE)

## Include portion and analyte data
TCGAbarcode(xbarcode, participant = TRUE, sample = TRUE, portion = TRUE)
```

## Sample selection

Based on lookup table values, the user can select certain sample types from a
vector of sample barcodes. Below we select "Primary Solid Tumors" from a vector
of barcodes, returning a logical vector identifying the matching samples.

```{r}
## Select primary solid tumors
TCGAsampleSelect(xbarcode, "01")

## Select blood derived normals
TCGAsampleSelect(xbarcode, "10")
```

### Primary tumors

We provide a `TCGAprimaryTumors` helper function to facilitate the selection
of primary tumor samples only:

```{r}
TCGAprimaryTumors(coad)
```

## `data.frame` representation of barcode

The straightforward `TCGAbiospec` function will take the information contained
in the TCGA barcode and display it in `data.frame` format with appropriate
column names.

```{r}
TCGAbiospec(xbarcode)
```

# OncoPrint - oncoPrintTCGA

We provide a convenience function that investigates metadata within
`curatedTCGAData` objects to present a plot of molecular alterations
within a paricular cancer. `MultiAssayExperiment` objects are required to
have an identifiable '*Mutation*' assay (using text search). The `variantCol`
argument identifies the mutation type column within the data.

**Note**. Functionality streamlined from the `ComplexHeatmap` package.

```{r}
oncoPrintTCGA(coad, matchassay = rag)
```

# Reference data

The `TCGAutils` package provides several helper datasets for working with TCGA barcodes.

## `sampleTypes`

As shown previously, the reference dataset `sampleTypes` defines sample codes
and their sample types (see `?sampleTypes` for source url).

```{r}
## Obtained previously
sampleCodes <- TCGAbarcode(xbarcode, participant = FALSE, sample = TRUE)

## Lookup table
head(sampleTypes)

## Match codes found in the barcode to the lookup table
sampleTypes[match(unique(substr(sampleCodes, 1L, 2L)), sampleTypes[["Code"]]), ]
```

Source: https://gdc.cancer.gov/resources-tcga-users/tcga-code-tables/sample-type-codes

## `clinicalNames` - Firehose pipeline clinical variables

`clinicalNames` is a list of the level 4 variable names (the most commonly used
clinical and pathological variables, with follow-ups merged) from each
`colData` datasets in `curatedTCGAData`. Shipped `curatedTCGAData`
`MultiAssayExperiment` objects merge additional levels 1-3 clinical,
pathological, and biospecimen data and contain many more variables than the ones
listed here.

```{r}
data("clinicalNames")

clinicalNames

lengths(clinicalNames)
```

# `sessionInfo`

```{r}
sessionInfo()
```
