---
title: "TCGAutils: Helper functions for working with TCGA datasets"
author: "Waldron Lab"
date: "`r format(Sys.time(), '%B %d, %Y')`"
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{TCGAutils Essentials}
  %\VignetteEncoding{UTF-8}
output:
  BiocStyle::html_document:
    number_sections: yes
    toc: true
---

```{r, echo=FALSE, warning=FALSE}
suppressPackageStartupMessages({
    library(TCGAutils)
    library(curatedTCGAData)
})
```

# Installation

```{r, eval=FALSE}
BiocInstaller::biocLite("TCGAutils")
```

Load some packages for the vignette:

```{r, eval=FALSE}
library(TCGAutils)
library(curatedTCGAData)
```

# Overview

The `TCGAutils` package provides some utility functions for working with
data from `The Cancer Genome Atlas`. Some functions are geared towards working
with existing tools and infrastructure such as `MultiAssayExperiment` and
`curatedTCGAData`. Other functions are intended to work with flat data
structures such as `data.frame` or `DataFrame`.
